Implementation of a stable, high-power optical lattice for quantum gas microscopy

### 2017-06-19 LatticeOptics/latticeOpticsV4.tex
new version

% -*- mode: Latex -*-
% Time-stamp: "2015-02-18 17:55:40 sb"

@@

\title{Design and Implementation of Stable High Power Optical Lattices for Quantum Gas Microscopy}

\author{Li Microscope Collective}
\author{M. Greiner}
\affiliation{
  Department of Physics, Harvard University,
  Cambridge, Massachusetts, 02138, USA}

\date{\today}
\begin{abstract}
We describe the design of a stable high power 1064 nm laser for use in optical lattice experiments. The system is based on a high quality mephisto NPRO seeding an array of four heavily modified fiber amplifiers. The intensity of every beam is stabilized with a low noise, nonlinear feedback system, and the waist can be smoothly controlled. 
\end{abstract}
\maketitle

\section{Introduction}
Ultracold atoms in optical lattices have become a powerful platform for experiments ranging from investigations of strongly correlated systems, quantum state engineering, quantum computing architectures, and precision metrology.
Further, the recent advent of quantum gas microscopy has enabled studies that exhibit unprecedented control over few atom systems at extremely low energy scales.
However, this rapid progress places ever more stringent technical constraints on the lasers used to trap and manipulate the atomic systems.
For example, heating due to the intensity noise of any laser cannot be allowed to approach the longest thermalization timescales, lest the system fail to thermalize~\cite{Savard1997}. 
Since thermalization timescales generally grow with falling temperature, this complicates attaining low temperatures.
Similarly, experiments that rely on a small number of atoms held by multiple traps places stringent requirements on the relative positional stability of the traps, to ensure that the operations performed by the traps have adequate fidelity.

In this note, we describe a high power optical lattice system that exhibits levels of intensity noise below $-120\,$dBc between $1\,$kHz and $3\,$MHz.
It also exhibits positional stability of the lattice phase at the $<0.5\,\mu$m level and of the harmonic confinement at the few site level over the course of hours.

The architecture of the system is based on a single low-power, low-noise seed laser (Innolight Mephisto) that supplies light to an array of four high-power fiber amplifiers (one for a transport beam~\cite{Huber2014}, and one each for the $X,Y,Z$ lattices)\cite{Parsons2015}.
Each fiber amplifier (Nufern NuAmp SUB1174-34) provides approximately 30 W, required to attain sufficient trap frequencies.
The fiber amplifier output is controlled by two feedback loops which stabilize the light in the low power and high power regimes.
After stabilization, the laser is mode-shaped to produce a desired waist when focused in the atom plane. 
Just before the beam is launched toward the atoms, a small sample of it is imaged by a camera to monitor the pointing.

To avoid interference effects among the lattice axes, AOM's are used to detune the fiber amplifier seeds away from each other.
These amplifiers can amplify 150 mW of seed light to up to approximately 45 W at the expense of dramatically raised levels of noise. 
The noise stems partially from stimulated Brillouin scattering, which is intrinsic, and several technical sources, which are not.

The NuAmp suffers from several problems: it is controlled via a USB interface, which is suspected to cause ground loops which add noise and interference to the analog systems, as seen in figure \ref{fig:noises}.
We solve this problem by replacing the USB control board with a custom solution that is compatible with our custom control system and that carefully avoids ground loops. 
There are two power supplies in the fiber amplifier - a general purpose supply for the control electronics and first amplification stage and a high power current supply for the second stage. 
Despite the presence of a water-cooled cold-plate to cool the optical fiber and pump diodes, the high power supply is air cooled using a fan approximately $20$ cm away from the gain fiber. 
This fan is suspected to lead to increased noise at acoustic frequencies.
In order to reduce acoustic coupling and reduce switching power supply noise, both power supplies were removed from the fiber amplifier chassis, and the low power supply was replaced by a linear PSU (Acopian A24MT350M). 
Further, a high current line filter (MPE DS26387) was added to the high-power PSU (Lumina LDD-600-60-10-5VP/M).
Combined, these changes serve to suppress many of the noise spurs measured in the RIN of the fiber amplifier, as seen in figure \ref{fig:noises}. 
The measurements in parts (a)-(d) were carried out on a single fiber amplifier, prior to and after the modifications. The initial and final noise spectra were verified to be consistent between amplifiers coming from the same manufacturing batch.
The figure is based on fiber amplifiers from a relatively new batch, which, even in its unmodified configuration does not exhibit a large switching supply spur. 
This is not the case for certain older batches, where the spur is approximately 30 dB higher than the background.

@@

This is possible because the pump diodes are fixed directly to the water-cooling plate without additional temperature regulation.
Adding a Peltier TEC and regulating to the optimal wavelength is possible, but is technically difficult as it requires large cooling power of $\approx100$ W per each of the three pump diodes and carries the associated complexity of three feedback loops.
We use a single, dedicated heat exchanger (TermoTek P21518970) to supply cooling water to an array of four fiber amplifiers and tune the temperature such that the average power supplied by the amplifiers supplying the $X$ and $Y$ lattices is optimized, while simultaneously keeping the temperature of the amplifier well within a safe range.
Experimentally, we have found that more power can be gained by running the amplifiers at hotter temperature, but we have avoided doing this in an effort to increase the lifetime.
We have experimentally found that there is variation in the sensitivity to the cooling water temperature between fiber amplifiers.

\begin{figure}
  \begin{center}
    \includegraphics{fig/fig3_combined.pdf}
    \caption{\textbf{Laser noise of the lattice system.} (a) Shows the laser noise of unmodified Nufern fiber amplifiers and the Mephisto seed. (b) Shows the open (OL) and closed (CL) loop behavior of the high power feedback loop. (c) Shows open loop behavior of the low power feedback loop after making the modifications. (d) Closed loop behavior of the loop system without additional filtering, measured on the same setup as (c).}\label{fig:noises}
  \end{center}
\end{figure}

\subsection{Mode and beam shaping}

\begin{figure}
  \begin{center}
    \includegraphics{fig/optical_layout_wide.pdf}
    \caption{\textbf{Optical layout of the optical lattice system for beam shaping and modulation.}}\label{fig:optical_layout}
  \end{center}
\end{figure}

@@

\subsection{High power feedback}

@@

After the isolation stage, we use the sequence of optical elements shown in figure \ref{fig:optical_layout}, consisting of a polarizer, half-wave plate (HWP), Berek compensator mounted on a precision galvo (Thorlabs GVS002, or Camtech 8320K, depending on beam path) half-wave plate, and polarizer. 
The Berek compensator is a simple $z$-cut quartz plate coated to be anti-reflecting at 1064 nm. 
By tilting the plate about its vertical axis, the extraordinary axis is mixed into the propagation of the beam, leading to a tilt-dependent bi-refringence that can be tuned from a zero-wave plate past a half-wave plate.
Combined with the subsequent polarization optics, rotation of the Berek compensator changes the transmitted power with perfect contrast and transmitted power as a function of the tilt angle is shown in figure \ref{fig:berek}.

Full modulation contrast is achieved when the fast axes of the waveplates around the Berek compensator $\theta$ are at $22.5^\circ$ to the rotation axis of the Berek compensator.
In that case, if the Berek compensator does not change the polarization, the light is fully transmitted. 
If it functions as a half-wave plate, the light is fully rejected, with a smooth crossover between the two cases as the Berek angle, $\phi$, is adjusted.
This contrast can be reduced by varying $\theta$, to the limiting case where $\theta=0$, where the rotation of the Berek compensator does nothing, since the polarization of the light is parallel to the rotation axis of the quartz.

\begin{figure}
  \begin{center}
    \includegraphics{fig/step_response.pdf}
    \caption{\textbf{Berek compensator based feedback loop} performance showing the impulse response (a) at relatively low powers and the corresponding Bode plots (b-c).}\label{fig:berek_step_response}
  \end{center}
\end{figure}
A key point is that, as seen in figure \ref{fig:berek}, if the two waveplates are rotated together, the maximum transmitted power is fixed, while the minimum can be raised.
The waveplates are detuned such that the maximum power needed in \textit{interaction} mode is the minimum power transmitted by the Berek compensator system, strongly suppressing fluctuations due to angle positioning noise of the Berek compensator. 
Naturally, rotation of a parallel plate in the beam path induces a shift of the beam, but this is acceptable because the Berek compensator is only active during the imaging phase of the experiment, when the position of the underlying harmonic trap is largely irrelevant.
Fortunately, the spatial phase of the optical lattice is set by the retro-reflector mirror, which does not move, and is not affected by the rotation of the Berek compensator.
This system functions as the actuator arm of a feedback loop, where the power of the beam is measured using a low noise photodiode. This feedback loop regulates the lattice between $1$ W of optical power and its maximum value (which depends on the amplifier, but is $\approx20$ W).
The step response and corresponding Bode plots~\cite{Bechhoefer2005} are shown in figure \ref{fig:berek_step_response}.
A polarization rotation by $2 \pi$ is produced by approximately $3^\circ$ of axial rotation of the quartz plate. 
Since we never need to rotate the plate outside this range, the position is constrained to this window in software.

\subsection{Low power feedback}
%The high power feedback used in the \textit{imaging} phase of the experiment and is exceedingly robust
%, and allows the user to raise the power and ensure an adequate Lamb-Dicke parameter\cite{Parsons2015}. 
The high power feedback used in the \textit{imaging} phase of the experiment is exceedingly robust because the heating rate from scattered resonant imaging light far exceeds any technical heating from the loop.
Even if the loop was designed by your worst enemy\footnote{I have received a number of comments about this particular statement. Yes, it was a hyperbole, and I concur - most likely readers of this thesis could design a system that would result in a heating rate exceeding the cooling, but you would definitely have to be malicious to do so.}, the heating would be entirely compensated by the simultaneous Raman cooling~\cite{Parsons2015}.

@@

\subsubsection{Phase-locked loop (PLL)}
The local oscillator (LO) RF source is a custom-built printed circuit board (PCB) implementing a phase locked loop (PLL) based on the ADF4002 IC and Crystek's CVCO55CL-0060-0110 voltage controlled oscillator (VCO), preamplified with a low noise amplifier (Minicircuits PSA4-5043+\footnote{This part has a plastic case which melts slightly if placed in a reflow oven, but still works as specified.}). 
To ensure frequency stability (which translates directly into beam pointing of the laser), the PLL is locked to the $10$ MHz clock distributed around the lab.
In addition to PLL functionality, the board also contains two, digitally-controllable, high-isolation RF switches, giving a total isolation exceeding $90$ dB.

The spectrum and relative intensity noise of the PLL can be found in figure \ref{fig:pll}. 
Vitally, the spectrum of the PLL exhibits phase noise comparable to that of typical VCO\footnote{Strictly speaking, a $-90$ dB phase noise is worse than specified. 
However, the measurement was carried out in ``peak detection'' mode on the spectrum analyzer, which artificially inflates the numbers by approximately 10 dB. 
This happens because the power at multiple frequency bands is measured within each frequency step on this plot, and in ``peak detection'', the highest is reported (all noise measurements in this work were carried out in ``RMS detection'', which is more suited for that purpose). The other reason that the phase noise appears worse than specified is that the dynamic range of the spectrum analyzer used is 100 dB, and we approach close to that limit.}, without the slow frequency drift characteristic to these devices.
Elimination of the carrier drift is worth the added phase noise because any drift of the frequency affects the steering of the laser beam passing through the AOM, which is entirely unacceptable since it would misalign the optical lattice.

@@

\begin{figure}
  \begin{center}
    \includegraphics[scale=0.1]{fig/cut_photodiode.jpg}
    \caption{\textbf{Photodiode photo} showing that the ``can'' of the photodiode has been cut, to prevent etalon-like behavior.}\label{fig:cut_pd}
  \end{center}
\end{figure}

Fast, low-noise photodiodes are required both to provide the monitoring arm of the feedback loop and to characterize the RIN of the system.
An ideal device would have a bandwidth up to approximately 3 MHz\footnote{The maximum attainable trap frequencies in the system are around 1.5 MHz, and parametric heating is driven by $2\times$ the trap frequency}, with a noise floor below approximately -150 dBc over that range.
To approach these specifications, we use a custom designed PCB that implements a simple trans-impedance amplifier~\cite{Graeme1995} in a convenient physical package, providing a bandwidths in the 0.1-10 MHz range, depending on the trans-impedance gain set by a resistor. 
The photodiode can be safely operated up to incident powers of approximately 10 mW, which sets a -164 dBc/Hz shot noise limit on the RIN that can be detected. 
In practice, we typically operate the photodiodes with $\approx 5$ mW of incident power.
A photodiode with a bandwidth (BW) of 6 MHz is used to characterize the system, while photodiodes with a bandwidth of 330 kHz (and correspondingly higher gain) are used to perform the feedback.

@@

Although it is an essential part of the system, the loop filter is also one of its major limitations. Figure \ref{fig:noises} shows the noise performance of the output of the fiber amplifiers in open loop as compared to the noise in closed loop. Although noise near DC is lowered (as expected from a closed loop system), noise outside the loop bandwidth is increased in the range of $10$ to $450$ kHz as seen in subfigures (c) and (d).
This can be explained by the Johnson noise~\cite{Horowitz2015} on the $1$ k$\Omega$ input resistors, amplified by the amplification stages of the loop filter\footnote{This seems like a small quantity (-174 dBm), but there are several levels of amplification, and the voltage supplied to the mixer is small ($\approx 0.1 V$), all of which works against the experimentalist.}. 
The sharp drop off at the $450$ kHz point is due to the low-pass filter centered there, placed on the output of the loop filter.
When this was determined to be a problem, and we decided that a bandwidth of higher than $10$ kHz was not required in the near term, a first order low pass with a $3$ dB frequency of $10$ kHz, was also added, leading to the noise spectrum seen in subfigure (b), as compared to (a) in figure \ref{fig:low_pass_noises}\footnote{The fiber amplifier used to take this data was different from the one used for \ref{fig:noises} different fiber amplifiers, which explains the small discrepancies.}.

\begin{figure}
  \begin{center}
    \includegraphics{fig/low_pass_filters.pdf}
    \caption{\textbf{Laser noise of the lattice system with and without additional low pass filters} (a) Closed loop laser made a fiber amplifier (different than the one in figure \ref{fig:noises}. (b) Noise on the same system as (a), but with extra filters.}\label{fig:low_pass_noises}
  \end{center}
\end{figure}

@@

\section{Beam monitoring and alignment}

Improved spatial resolution and control have given experimentalists the ability to use relatively few atoms~\cite{Preiss2015, Choi2016, Mazurenko2016} that are simultaneously addressed by multiple beams and held by multiple traps.
Such experiments place stringent bounds on the requisite stability.

The absolute positions of lattice sites cannot be allowed to change by more than approximately $10\%$ from experimental shot to shot.
Fortunately, this is relatively straightforward to ensure via careful design of the optical assembly, since the phase of a retro-reflected lattice is set by the retro-reflecting mirror, which can be close to the position of the atoms and be made very stable~\cite{Huber2014}. 
We have characterized the phase stability of the optical lattice by looking at a long sequence of densely populated images, and found that the phase varies by much less than one lattice site over the course of hours.

The position of the atomic cloud, however, is not determined by the phase of the lattice, but by the underlying harmonic confinement, which, in turn, is determined by the positions of the ingoing and retro-reflected beams. 
In experiments where the entire cloud is addressed by beams auxiliary to the lattice (such as in this work, \cite{Mazurenko2016} or \cite{Choi2016}), drift or fluctuation of the lattice position can pose a severe limitation.

The required positional stability is achieved through a combination of passive stability and active monitoring. 
Passive stability is ensured by using stable optomechanics (from the Thorlabs Polaris line), mounted at a low ($2$ in) beam height relative to the floating optical table.
The beam is enclosed with aluminum tubes throughout its path and the entire setup is enclosed, to suppress air currents.

Passive stability alone has been found to be insufficient to ensure smooth operation over many hours because external factors can lead to a slow drift in the pointing of the laser beams. 
To address this concern, each lattice axis is fitted with a remotely actuated mirror connected to the control software.
The actuators are Newport brand ``Picomotors'' (model 8302), fitted into a customized $2$ in mirror mount.

>>>>

@@

\title{Implementation of a stable, high-power optical lattice for quantum gas microscopy}

\author{A.M., S.B., F.H., M.P., C.C., G.J., D.G., M.G.}
\author{M. Greiner}
\affiliation{
  Department of Physics, Harvard University,
  Cambridge, Massachusetts, 02138, USA}

\date{\today}
\begin{abstract}
We describe the design of a stable high power $1064\,$nm laser system for use in optical lattice experiments. The system is based on a low noise seed laser amplified with an array of four heavily modified \textit{Nufern} fiber amplifiers. The intensity of every beam is stabilized with a low noise, nonlinear feedback loop. The positional stability of the system is characterized and monitored in real time.  
\end{abstract}
\maketitle

Ultracold atoms trapped in an optical lattice has become a powerful platform for experiments ranging from investigations of strongly correlated systems, quantum state engineering, quantum computing architectures and precision metrology\cite{Ye}.
Further, the recent advent of quantum gas microscopy has enabled studies that exhibit unprecedented control over few atom systems at extremely low energy scales.
However, this rapid progress places ever more stringent technical constraints on the lasers used to trap and manipulate the atomic systems.
For example, heating due to the intensity noise of any laser cannot be allowed to approach the longest thermalization timescales, lest the system fail to thermalize~\cite{Savard1997}. 
Since thermalization timescales generally grow with falling temperature, this complicates attaining low temperatures.
Similarly, experiments that rely on a small number of atoms held by multiple traps places stringent requirements on the relative positional stability of the traps, to ensure that the operations performed by the traps have adequate fidelity.

To address these concerns, we discuss a high power optical lattice system that exhibits levels of intensity noise below $-120\,$dBc between $1\,$kHz and $3\,$MHz.
The positional stability of the lattice phase is at the $<0.5\mu$m level and the stability of the harmonic confinement is at the few site level, measured over the course of hours.

\begin{figure}
  \begin{center}
    \includegraphics{fig/figure1/figure1.pdf}
    \caption{\textbf{High level overview} showing the major components of the lattice laser feedback system.}
  \end{center}
\end{figure}

The architecture is based on a single low-power, low-noise seed laser that supplies light to an array of four high-power fiber amplifiers (one for a transport beam~\cite{Huber2014}, and one each for the $X,Y,Z$ lattices).
These amplifiers generate enough light to perform site resolved imaging~\cite{Parsons2016}, but also substantially increase the level of noise and degrade the mode shape. 
The fiber amplifier output is controlled by two feedback loops, which stabilize the light in the low power and high power regimes.
After stabilization, the laser is mode-shaped to produce a desired waist when focused in the atom plane. 
Just before the beam is launched toward the atoms, a small sample of it is imaged by a camera to monitor the pointing.

\section{Laser system}
The lattice laser system is seeded by an Innolight Mephisto laser, which provides approximately 2 W of 1064 nm light. 
This commercial laser utilizes a non-planar ring oscillator (NPRO) design that ensures superb phase and intensity stability. This source is split into four fibers (Thorlabs PM980-XP) using polarizing beam splitters (PBS).
To avoid interference of the lattice beams, two of the beam paths pass through acousto-optic modulators (AOM), which detunes the lattice frequencies from each other. Figure \ref{fig:noises} shows the noise performance of only this subsystem.

Quantum gas microscopy of \Li{} requires single site trap frequencies of approximately 1 MHz~\cite{Parsons2015} in order to be in the Lamb-Dicke regime, which is necessary to perform Raman sideband cooling.
Such trap frequencies may only be attained using tightly focused high power beams, which requires that the seed power be amplified to approximately 30 W per lattice axis, which is done using Nufern Fiber amplifiers (NuAmp SUB1174-34).
These amplifiers can amplify 150 mW of seed light to up to approximately 45 W. They use a length of doped fiber pumped with 808 nm light from fiber coupled diode sources to amplify 1064 nm light. 
The amplification is performed using a two stage amplifier design, where the stages are controlled using separate power supplies and electronics.
The high optical powers supplied by the fiber amplifiers raise two engineering difficulties:
\begin{enumerate}
  \item The fiber amplification process is susceptible to many technical noise sources, such as acoustic coupling of the fiber to the environment and electrical design of the amplifier, 
  \item Stimulates Brillouin scattering (SBS) places a hard lower bound on the noise floor~\cite{Agrawal2013}. 
\end{enumerate}

The NuAmp suffers from several problems: it is controlled via a USB interface, which is suspected to cause ground loops which add noise and interference to the analog systems, as seen in figure \ref{fig:noises}.
We solve this problem by replacing the USB control board with a custom-designed, Ethernet based solution that is compatible with our custom control system and that carefully avoids ground loops. 
There are two power supplies in the fiber amplifier - a general purpose supply for the control electronics and first amplification stage and a high power current supply for the second stage. 
Despite the presence of a water-cooled cold-plate to cool the optical fiber and pump diodes, the high power supply is air cooled using a fan approximately $20$ cm away from the gain fiber. 
This fan is suspected to lead to increased noise at acoustic frequencies.
In order to reduce acoustic coupling and reduce switching power supply noise, both power supplies were removed from the fiber amplifier chassis, and the low power supply was replaced by a linear PSU (Acopian A24MT350M). 
Further, a high current line filter (MPE DS26387) was added to the high-power PSU (Lumina LDD-600-60-10-5VP/M).
Combined, these changes serve to suppress many of the noise spurs measured in the RIN of the fiber amplifier, as seen in figure \ref{fig:noises}. 
The measurements in parts (a)-(d) were carried out on a single fiber amplifier, prior to and after the modifications. The initial and final noise spectra were verified to be consistent between amplifiers coming from the same manufacturing batch.
The figure is based on fiber amplifiers from a relatively new batch, which, even in its unmodified configuration does not exhibit a large switching supply spur. 
This is not the case for certain older batches, where the spur is approximately 30 dB higher than the background.

@@

This is possible because the pump diodes are fixed directly to the water-cooling plate without additional temperature regulation.
Adding a Peltier TEC and regulating to the optimal wavelength is possible, but is technically difficult as it requires large cooling power of $\approx100$ W per each of the three pump diodes and carries the associated complexity of three feedback loops.
We use a single, dedicated heat exchanger (TermoTek P21518970) to supply cooling water to an array of four fiber amplifiers and tune the temperature such that the average power supplied by the amplifiers supplying the $X$ and $Y$ lattices is optimized, while simultaneously keeping the temperature of the amplifier well within a safe range.
Experimentally, we have found that more power can be gained by running the amplifiers at hotter temperature, but we have avoided doing this in an effort to increase the lifetime.
We have experimentally found that there is variation in the sensitivity to the cooling water temperature between fiber amplifiers. 
%Figure\ref{fig:pow_and_noise}c. shows the responses of the fiber amplifiers currently in the system.

\begin{figure*}
  \begin{center}
    \includegraphics{fig/fig3_combined.pdf}
    \caption{\textbf{Laser noise of the lattice system.} (a) Shows the laser noise of unmodified Nufern fiber amplifiers and the Mephisto seed. (b) Shows the open (OL) and closed (CL) loop behavior of the high power feedback loop. (c) Shows open loop behavior of the low power feedback loop after making the modifications. (d) Closed loop behavior of the loop system without additional filtering, measured on the same setup as (c).}\label{fig:noises}
  \end{center}
\end{figure*}

\begin{figure*}
  \begin{center}
    \includegraphics{fig/optical_layout_wide.pdf}
    \caption{\textbf{Optical layout of the optical lattice system for beam shaping and modulation.}}\label{fig:optical_layout}
  \end{center}
\end{figure*}

@@

\section{Feedback System}

@@

After the isolation stage, we use the sequence of optical elements shown in figure \ref{fig:optical_layout}, consisting of a polarizer, half-wave plate (HWP), Berek compensator mounted on a precision galvo (Thorlabs GVS002, or Camtech 8320K, depending on beam path) half-wave plate, and polarizer. 
The Berek compensator is a simple $z$-cut quartz plate coated to be anti-reflecting at 1064 nm. 
By tilting the plate about its vertical axis, the extraordinary axis is mixed into the propagation of the beam, leading to a tilt-dependent bi-refringence that can be tuned from a zero-wave plate past a half-wave plate.
Combined with the subsequent polarization optics, rotation of the Berek compensator changes the transmitted power with perfect contrast and transmitted power as a function of the tilt angle is shown in figure \ref{fig:berek}.
\begin{figure}
  \begin{center}
    \includegraphics{fig/berek_sim.pdf}
    \caption{\textbf{Berek compensator} performance as a function of the Berek tilt angle $\phi$ and waveplate tilt angles.}\label{fig:berek}
  \end{center}
\end{figure}
Full modulation contrast is achieved when the fast axes of the waveplates around the Berek compensator $\theta$ are at $22.5^\circ$ to the rotation axis of the Berek compensator.
In that case, if the Berek compensator does not change the polarization, the light is fully transmitted. 
If it functions as a half-wave plate, the light is fully rejected, with a smooth crossover between the two cases as the Berek angle, $\phi$, is adjusted.
This contrast can be reduced by varying $\theta$, to the limiting case where $\theta=0$, where the rotation of the Berek compensator does nothing, since the polarization of the light is parallel to the rotation axis of the quartz.

\begin{figure*}
  \begin{center}
    \includegraphics{fig/step_response.pdf}
    \caption{\textbf{Berek compensator based feedback loop} performance showing the impulse response (a) at relatively low powers and the corresponding Bode plots (b-c).}\label{fig:berek_step_response}
  \end{center}
\end{figure*}
A key point is that, as seen in figure \ref{fig:berek}, if the two waveplates are rotated together, the maximum transmitted power is fixed, while the minimum can be raised.
The waveplates are detuned such that the maximum power needed in \textit{interaction} mode is the minimum power transmitted by the Berek compensator system, strongly suppressing fluctuations due to angle positioning noise of the Berek compensator. 
Naturally, rotation of a parallel plate in the beam path induces a shift of the beam, but this is acceptable because the Berek compensator is only active during the imaging phase of the experiment, when the position of the underlying harmonic trap is largely irrelevant.
Fortunately, the spatial phase of the optical lattice is set by the retro-reflector mirror, which does not move, and is not affected by the rotation of the Berek compensator.
This system functions as the actuator arm of a feedback loop, where the power of the beam is measured using a low noise photodiode. This feedback loop regulates the lattice between $1$ W of optical power and its maximum value (which depends on the amplifier, but is $\approx20$ W).
The step response and corresponding Bode plots~\cite{Bechhoefer2005} are shown in figure \ref{fig:berek_step_response}.
A polarization rotation by $2 \pi$ is produced by approximately $3^\circ$ of axial rotation of the quartz plate. 
Since we never need to rotate the plate outside this range, the position is constrained to this window in software.

\subsection{Low power feedback}
%The high power feedback used in the \textit{imaging} phase of the experiment and is exceedingly robust
%, and allows the user to raise the power and ensure an adequate Lamb-Dicke parameter\cite{Parsons2015}. 
The high power feedback used in the \textit{imaging} phase of the experiment is exceedingly robust because the heating rate from scattered resonant imaging light far exceeds any technical heating from the loop.
Even if the loop was exceedingly noisy, the heating would be entirely compensated by the simultaneous Raman cooling~\cite{Parsons2015}.

@@

\subsubsection{Phase-locked loop (PLL)}
The local oscillator (LO) RF source is a custom-built printed circuit board (PCB) implementing a phase locked loop (PLL) based on the ADF4002 IC and Crystek's CVCO55CL-0060-0110 voltage controlled oscillator (VCO), preamplified with a low noise amplifier (Minicircuits PSA4-5043+). 
To ensure frequency stability (which translates directly into beam pointing of the laser), the PLL is locked to the $10$ MHz clock distributed around the lab.
In addition to PLL functionality, the board also contains two, digitally-controllable, high-isolation RF switches, giving a total isolation exceeding $90$ dB.

The spectrum and relative intensity noise of the PLL can be found in figure \ref{fig:pll}. 
Vitally, the spectrum of the PLL exhibits phase noise comparable to that of typical VCO, without the slow frequency drift characteristic to these devices. 
Elimination of the carrier drift is worth the added phase noise because any drift of the frequency affects the steering of the laser beam passing through the AOM, which is entirely unacceptable since it would misalign the optical lattice.

@@

\begin{figure}
  \begin{center}
    \includegraphics[width=\columnwidth]{fig/cut_photodiode.jpg}
    \caption{\textbf{Photodiode photo} showing that the ``can'' of the photodiode has been cut, to prevent etalon-like behavior.}\label{fig:cut_pd}
  \end{center}
\end{figure}

Fast, low-noise photodiodes are required both to provide the monitoring arm of the feedback loop and to characterize the RIN of the system.
An ideal device would have a bandwidth up to approximately 3 MHz (since trap frequencies in the system are $\approx1.5\,$MHz), with a noise floor below approximately -150 dBc over that range.
To approach these specifications, we use a custom designed PCB that implements a simple trans-impedance amplifier~\cite{Graeme1995} in a convenient physical package, providing a bandwidths in the 0.1-10 MHz range, depending on the trans-impedance gain set by a resistor. 
The photodiode can be safely operated up to incident powers of approximately 10 mW, which sets a -164 dBc/Hz shot noise limit on the RIN that can be detected. 
In practice, we typically operate the photodiodes with $\approx 5$ mW of incident power.
A photodiode with a bandwidth (BW) of 6 MHz is used to characterize the system, while photodiodes with a bandwidth of 330 kHz (and correspondingly higher gain) are used to perform the feedback.

@@

Although it is an essential part of the system, the loop filter is also one of its major limitations. Figure \ref{fig:noises} shows the noise performance of the output of the fiber amplifiers in open loop as compared to the noise in closed loop. Although noise near DC is lowered (as expected from a closed loop system), noise outside the loop bandwidth is increased in the range of $10$ to $450$ kHz as seen in subfigures (c) and (d).
This can be explained by the Johnson noise~\cite{Horowitz2015} on the $1$ k$\Omega$ input resistors, amplified by the amplification stages of the loop filter.
The sharp drop off at the $450$ kHz point is due to the low-pass filter centered there, placed on the output of the loop filter.
When this was determined to be a problem, and we decided that a bandwidth of higher than $10$ kHz was not required in the foreseeable future, a first order low pass with a $3$ dB frequency of $10$ kHz was  added, leading to the noise spectrum seen in subfigure (b), as compared to (a) in figure \ref{fig:low_pass_noises} (the fiber amplifier used to take this data was different from the one used for \ref{fig:noises} different fiber amplifiers, which explains the small discrepancies).

\begin{figure*}
  \begin{center}
    \includegraphics{fig/low_pass_filters.pdf}
    \caption{\textbf{Laser noise of the lattice system with and without additional low pass filters} (a) Closed loop laser made a fiber amplifier (different than the one in figure \ref{fig:noises}. (b) Noise on the same system as (a), but with extra filters.}\label{fig:low_pass_noises}
  \end{center}
\end{figure*}

@@

\subsection{Heating rates}
We have used the band mapping technique, described in \ref{sec:bloch}, to characterize the heating rates in the lattice at different powers~\cite{Blatt2015}.
To do this, the cloud was held in a given lattice for a varied period of time, before band mapping was performed.
Band mapping directly reveals the band occupations, and the lifetime of atoms in the ground band is closely related to the temperature.
The results, shown in figure \ref{fig:heating_rates} revealed that lifetimes as high as tens of seconds can be achieved, but that the heating rates are consistently dominated by parametric heating induced transitions.
This analysis was performed prior to adding extra 10 kHz filters to the system, and later measurements have revealed that heating rates in the $0.01-0.5$ W are approximately a factor of $2-5\times$ lower.

\section{Beam monitoring and alignment}
Improved spatial resolution and control have given experimentalists the ability to use relatively few atoms~\cite{Preiss2015, Choi2016, Mazurenko2016} that are simultaneously addressed by multiple beams and held by multiple traps.
Such experiments place stringent bounds on the requisite stability.

The absolute positions of lattice sites cannot be allowed to change by more than a small fraction ($\approx 10\%$) from experimental shot to shot.
Conveniently, this is relatively straightforward to ensure via careful design of the optical assembly, since the phase of a retro-reflected lattice is set by the retro-reflecting mirror, which can be close to the position of the atoms and be made very stable~\cite{Huber2014}. 
We have characterized the phase stability of the optical lattice by looking at a long sequence of densely populated images, and found that the phase varies by much less than one lattice site over the course of hours, as seen in figure \ref{fig:averaged_mott}.

\begin{figure}
  \begin{center}
    \includegraphics[width=\columnwidth]{fig/averaged_mott.png}
    \caption{\textbf{Averaged images of Mott insulators}. A series of Mott insulators was created over the course of approximately one hour, and the average image is shown\cite{Greif2016}. If the lattice fluctuated across more than a small fraction of a lattice site, the modulation contrast would be quickly washed out. The absence of this effect means that the phase of the optical lattice is very stable.}\label{fig:averaged_mott}
  \end{center}
\end{figure}

The position of the atomic cloud, however, is not determined by the phase of the lattice, but by the underlying harmonic confinement. 
In experiments where the entire cloud is addressed by beams auxiliary to the lattice (such as in this work, \cite{Mazurenko2016} or \cite{Choi2016}), drift or fluctuation of the lattice position can pose a severe limitation.

The required positional stability is achieved through a combination of passive stability and active monitoring. 
Passive stability is ensured by using stable optomechanics (largely from the Thorlabs Polaris line), mounted on short optical posts, ensuring a low ($2$ in) beam height relative to the optical table, which is floated to minimize vibrations.
In order to further improve stability, the beam is enclosed with $1$ in aluminum tubes throughout its path and the entire setup is enclosed by metal plates\footnote{all plates are sanded to produce a rough finish, producing diffuse rather than specular reflections.}.
To further isolate the experimental setup, the optical table is contained within an aluminum enclosure. As detailed in prior sections, care has been taken to minimize thermal lensing and drift in optics encountered by the lattice beam.

Passive stability alone has been found to be insufficient to ensure smooth operation over many hours. Changes in weather conditions, load on the water-cooling systems and other external effects can lead to a slow drift in the pointing of the laser beams. 
The situation is further complicated by the fact that opening the enclosure changes the thermal state of the system, inducing misalignment. 
For these reasons, each lattice axis is fitted with a remotely actuated mirror, controlled by the experimental control software over the Ethernet network, allowing for non-invasive, digital alignment of the system. 
The remote actuators are Newport brand ``Picomotors'' (model 8302), fitted into a customized $2$ in mirror mount, and work in a ``fire and forget'' manner, remaining passively stable when not in active use.

\begin{figure*}
  \begin{center}
    \includegraphics{fig/pointing_figure.pdf}
    \caption{\textbf{Short term positional stability.} (a) A histogram of shot-to-shot step sizes along the transverse and longitudinal directions, with a Gaussian fit. (b) Correlations between the step size of subsequent shots. (c) Calibration data for the ``affine fit'' of the conversion from camera pixels to lattice dimentions. (d) Predicted location versus measured location along the transverse and longitudinal directions. (e,f) The measured position and predicted position of the atomic cloud.}\label{fig:stability}
  \end{center}
\end{figure*}

To prevent external influences from contaminating large data sets, we have implemented an independent monitoring system to verify appropriate pointing of both lattice laser beams and eliminate (via post-selection) experimental realizations when the system was in an undesired state.
A beam-sampler (Thorlabs BSF10-C) reflects a small portion of the beam shortly before it reaches the atoms and images it onto a camera located in the same plane as the atoms, producing a Gaussian image that is identical to that incident on the atoms.
Naturally, the position of the atoms is dictated by the light field of both the incoming and retro-reflected beams, while the monitoring camera only observes the position of the in-going beam.
To verify that the monitor camera could predict the true position of the atoms, the beam and the atoms were imaged simultaneously while a purposeful misalignment of the incoming beams was introduced.
This was done by adiabatically loading one of the lattice beams ensuring that the atoms settle to its center, prior to performing site-resolved imaging~\cite{Parsons2015}.
